Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2018-09-18 writeup/local_pca_paper.tex
resubmitting now

Population structure leads to systematic patterns in measures of mean relatedness
between individuals in large genomic datasets,
which are often discovered and visualized using dimension reduction techniques such as principal component analysis (PCA).
Mean relatedness
is an average of the relationships across locus-specific genealogical trees,
which can be strongly affected on intermediate genomic scales by linked selection and other factors.
We show how to use local principal components analysis to describe this 
meso-scale heterogeneity in patterns of relatedness,
and apply the method to genomic data from three species,
finding in each that the effect of population structure 
can vary substantially across only a few megabases.
In a global human dataset, localized heterogeneity is likely explained by polymorphic chromosomal inversions.
In a range-wide dataset of \textit{Medicago truncatula},
factors that produce heterogeneity are shared between chromosomes,
correlate with local gene density,
and may be caused by background selection or local adaptation.
In a dataset of primarily African \textit{Drosophila melanogaster}, large-scale heterogeneity across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection,
and after removing samples carrying inversions, remaining heterogeneity is correlated with recombination rate and gene density,
again suggesting a role for linked selection.
% The method is implemented as an R package, called lostruct.
The visualization method provides a flexible new way to discover biological drivers of genetic variation,
and its application to data highlights the strong effects that linked selection and chromosomal inversions
can have on observed patterns of genetic variation.

@@

We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
in the space defined by the top principal components.
However, 
we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
since rotations or reflections of these imply identical patterns of relatedness.
Instead, we compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs,
which is invariant under ordering of the PCs \revpoint{1}{2},
reflections, and rotations and yet contains all other information about the PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
Furthermore, to remove the effect of artifacts such as mutation rate variation,
we also rescale each approximate covariance matrix to be of similar size
(precisely, so that the underlying data matrix has trace norm equal to one).

To do this, define the $N \times k$ matrix $V(i)$ so that $V(i)_{\cdot \ell}$, 
the $\ell^\text{th}$ column of $V(i)$,
is equal to the $\ell^\text{th}$ principal component of the $i^\text{th}$ window,
multiplied by $( \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} )^{1/2}$,
where $\lambda_{\ell i}$ is the $\ell^\text{th}$ eigenvalue of the genetic covariance matrix.
% \begin{equation}
%     \begin{aligned}
%         V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} ,
%         \qquad
%         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} 
%     \end{aligned}
% \end{equation}
Then, the rescaled, rank $k$ approximate covariance matrix for the $i^\text{th}$ window is
\begin{align} \label{eqn:est_cov}
    M(i) &= \sum_{\ell=1}^k V(i)_{\cdot \ell} V(i)_{\cdot \ell}^T .
\end{align}
% \begin{align}
%     M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
% \end{align}
To measure the similarity of patterns of relatedness for the $i^\text{th}$ window and $j^\text{th}$ window,
we then use
Euclidean distance $D_{ij}$ between the matrices $M(i)$ and $M(j)$:
$D_{ij}^2 = \sum_{k\ell} ( M(i)_{k,\ell} - M(j)_{k,\ell} )^2$.

@@

We tested the method using two types of simulation.
First, to verify expected behavior, we simulated ``genomes'' as an independent sequence
of correlated Gaussian ``genotypes'', using a different covariance matrix
in the first quarter, middle half, and last quarter of the chromosome.
To verify robustness to missing data,
we ran the method after randomly dropping 50\% of the genotypes in the first half of the genome;
if the method is misled by missing data, then it will distinguish the two halves of the chromosome
rather than the segments having different covariance matrices.
Details are given in Appendix \ref{apx:gaussian_sims}.

To provide a realistic test,
we next used forwards-time, individual-based simulations, implemented using SLiM v3 \citep{haller2017flexible},
which are described in detail in Appendix \ref{apx:slim_sims}.
To provide realistic population structure for PCA to identify,
each simulation had at least 5,000 diploid individuals,   
living across a continuous square range,
with Gaussian dispersal and local density-dependent competition.
Each genome was modeled on human chromosome 7, which is $1.54 \times 10^8$bp long,
with an overall recombination rate of 1.6785 crossovers per chromosome per generation.
To improve speed, we used tskit \citep{kelleher2018efficient} to record tree sequences
and to add neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
Most simulations were neutral, but we also included linked selection, of two types.
First, % run_031486
we introduced selected mutations into two regions, 
which extended from 1/3 to 1/2 and from 5/6 to the end of the genome respectively.
These had selection coefficients from a Gamma distribution with shape 2 and mean 0.005
at a rate of $10^{-10}$ per bp, that were either beneficial (with probability $1/30$)
or deleterious (otherwise).
Second, % run_016913
to roughly model a recent expansion followed by local adaptation,
we introduced mutations in the same manner as above,
except that mutations were no longer unconditionally deleterious or beneficial:
each selection coefficient was multiplied by a factor depending on the spatial location
of the individual being evaluated, varying linearly from -1 at the left side of the range
to +1 at the right edge.
In all simulations, genome-wide PCA displayed a map of the population range, as expected.

@@

Simple non-population-based simulations with Gaussian ``genotypes''
showed that the method performs as expected, clearly separating regions of the genome
with different underlying covariance matrices
without being affected by extreme differences in amount of missing data
(Supplemental Figure \ref{fig:test_missing}).
This simulation also verifies insensitivity to ordering of top PCs,
since it was performed using a covariance matrix with the top two eigenvalues equal,  
so that the order of empirical eigenvectors (PCs) switches randomly.

Individual-based simulations using SLiM \citep{haller2017flexible} allowed us to test the effects
of recombination and mutation rate variation, as well as linked selection.
Results are shown in Supplemental Figures \ref{sfig:slim_sims1} and \ref{sfig:slim_sims2}.
As expected, varying recombination rate stepwise by a factor of 64
did not induce patterns in the MDS visualizations correlated with recombination rate.
\llabel{ll:test_recomb_rate}
Since varying mutation rate with a fixed recombination map
is equivalent to varying the recombination map and remapping windows,
this also indicates that the method is not misled by variation in mutation rate.
On the other hand,
a recombination map with hotspots
(the HapMap human female map for chromosome 7 \citep{hapmap2007second})
induced outliers at long regions of low recombination rate
(also as expected). \llabel{ll:hotspot_recomb}

@@

We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
The first column of \Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
of the matrix of dissimilarities between genomic windows:
in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
For each chromosome arm, the MDS visualization roughly resembles a triangle,
sometimes with additional points.
Since the relative position of each window in this plot shows the similarity between windows, 
this suggests that there are at least three extreme manifestations of population structure 
typified by windows found in the ``corners'' of the figure,
and that other windows' patterns of relatedness may be a mixture of those extremes. 
The next two columns of \Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
plotted against the window's position along the genome,
to show how the plot of the first column is laid out along the genome.
The patterns did not depend on the number of PCs used
(see \Figure \ref{fig:dmel_k5} for the same plot with $k=5$ PCs), \revpoint{1}{3}
and are only weakly correlated with variation in missingness 
(see \Figure \ref{sfig:dpgp_missingness}). \revpoint{2}{4}

@@

To help visualize how clustered windows with similar patterns of relatedness are along each chromosome arm, 
we selected three ``extreme'' windows in the MDS plot
and the 5\% of windows that are closest to it in the MDS coordinates,
then highlighted these windows' positions along the genome,
and created PCA plots for the windows, combined.
Representative plots are shown for three groups of windows on each chromosome arm
in \Figure \ref{fig:mds_allchr} (groups are shown in color),
and in Supplemental \Figure \ref{fig:pca_by_pop} (PCA plots).
The latter plots are quite different,
showing that genomic windows in different regions of the MDS plot 
indeed show quite different patterns of relatedness.

The most striking variation in patterns of relatedness turns out to be explained by
several large inversions that are polymorphic in these samples, 
discussed in \citet{corbett2012population} and \citet{langley2012genomic}.
% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
To depict this, \Figure \ref{fig:pca_by_inversion} shows
the PCA plots in \Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
Taking chromosome arm 2L as an example,
the two regions of similar, extreme patterns of relatedness
shown in green in the first row of \Figure \ref{fig:mds_allchr}
lie directly around the breakpoints of the inversion In(2L)t,
and the PCA plots in the first rows of \Figure \ref{fig:pca_by_inversion}
shows that patterns of relatedness here are mostly determined by inversion orientation.
The regions shown in purple on chromosome 2L lie near the centromere,
and have patterns of relatedness reflective of two axes of variation,
seen in \Figures \ref{fig:pca_by_pop} and \ref{fig:pca_by_inversion},
which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
(see \citet{lack2015drosophila} for more about admixture in this sample).
The regions shown in orange on chromosome 2L mostly lie inside the inversion,
and show patterns of relatedness that are a mixture between the other two,
as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
Similar results are found in other chromosome arms,
albeit complicated by the coexistence of more than one polymorphic inversion;
however, each breakpoint visibly affects patterns in the MDS coordinates
(see vertical lines in \Figure \ref{fig:mds_allchr}).

@@

To see how patterns of relatedness vary in the absence of polymorphic inversions,
we performed the same analyses after removing, for each chromosome arm,
any samples carrying inversions on that arm.
In the result, shown in Supplemental \Figure \ref{sfig:mds_allchr_noinversions},
the striking peaks associated with inversion breakpoints are gone,
and previously smaller-scale variation now dominates the MDS visualization.
For instance, the majority of the variation along 3L in \Figure \ref{fig:mds_allchr}
is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
there is also a relatively small dip on the right end of the arm (near the centromere).
In contrast, Supplemental \Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
remaining structure is dominated by the dip near the centromere.
Without inversions, variation in patterns of relatedness shown in the MDS plots
follows similar patterns to that previously seen in \textit{D.~melanogaster} recombination rate and diversity \citep{langley2012genomic,mackay2012drosophila}.
Indeed, correlations between the recombination rate in each window and the position on the first MDS coordinate are highly significant
(Spearman's $\rho=0.54$, $p<2 \times 10^{-16}$; \Figures \ref{fig:drosophila_recomb_rate} and \ref{sfig:drosophila_gene_density_correlations}).
This is consistent with the hypothesis that variation
is due to selection, since the strength of linked selection increases with local gene density, 
measured in units of recombination distance.
The number of genes -- measured as the number of transcription start and end sites within each window --
was not significantly correlated with MDS coordinate ($p=0.22$).

@@

The results were highly consistent across window sizes, window types (SNPs or bp),
and number of PCs, as shown in Table \ref{tab:param_cors}.
\revpoint{1}{3} \revpoint{1}{4} \revpoint{1}{5}
% We also found nearly identical results when choosing shorter windows of 1,000 SNPs
% (Supplemental Figure \ref{sfig:mds_medicago_allchr_1e3}; \revpoint{1}{4}
% or choosing windows of equal length in base pairs rather than SNPs. \revpoint{1}{5}
% Similarly, the results were not substantially changed
% when using weighted PCA to downweight the large group of Tunisian samples.

\begin{figure}
    \begin{center}
       \includegraphics{Fig6_Together_MDS_plot_chr3_final}
       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
    \end{center}
    \caption{
        \textbf{(top)}
       MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3,
       with corresponding PCA plots.
       Each point in the plot represents a window;
       the structure revealed by the MDS plot is strongly clustered along the chromosome,
       with windows in the upper-right corner of the MDS plot (colored red) clustered around the centromere,
       windows in the upper-left corner (purple) furthest from the centromere,
       and the remaining corner (green) intermediate.
       Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:mds_medicago_allchr}.
       \textbf{(bottom)}
        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in the top figure.
        Each point corresponds to a sample, colored by country of origin.
        Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:pca_peaks_medicago_allchr}.
       \label{fig:mds12_medicago}
       \label{fig:pca_medicago}
    }
\end{figure}

@@

\paragraph{Chromosomal inversions}
A major driver of variation in patterns of relatedness in two datasets we examined are inversions.
This may be common,
but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
however, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
or simply regions of reduced recombination
without additional lines of evidence.
Inversions show up in our method because across the inverted region,
most gene trees share a common split that dates back to the origin of the inversion.
However, in many applications, inversions are a nuisance.
For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
by regressing out the effect of linked SNPs on each other.
Removing samples with the less common orientation of each inversion reduced,
but did not eliminate, the signal of inversions
seen in the \textit{Drosophila melanogaster} dataset,
demonstrating that the genomic effects of transiently polymorphic inversions
may outlast the inversions themselves.

@@

Another possible contributor is recent admixture between previously separated populations,
the effects of which were not uniform across the genome due to selection. \revpoint{2}{6}
For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
is due to selection against Neanderthal genes, leading to greater introgression in regions of lower gene density
\citep{harris2016genetic,juric2016strength}.
African \textit{Drosophila melanogaster} are known to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
if selection regularly favors genes from one origin,
this could lead to substantial variation in patterns of relatedness correlated with local gene density.

There has been substantial debate over the relative impacts of different forms of selection
(e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining}). \revpoint{2}{16}
These have been difficult to disentangle in part because for the most part
theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
It may be possible to design more powerful statistics that make stronger use of spatial information.

@@

\paragraph{Parameter choices}
There are several choices in the method that may in principle affect the results.
As with whole-genome PCA,
the choice of samples is important,
as variation not strongly represented in the sample will not be discovered.
The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
but downweighting may be a better option that does not discard data
(and here we present a method to do this).
Next, the choice of window size may be important,
although in our applications results were not sensitive to this,
indicating that we can see variation on a sufficiently fine scale.
Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in \Figure \ref{fig:diagram}),
along with the method used to discover common structure,
will affect results.
We used MDS, applied to either each chromosome separately or to the entire genome;
for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosomes,
but genome-wide, their signal is obscured by the numerous other signals of comparable strength.

@@

We first created ``genotypes'' in this way with fifty individuals from each of the three populations;
running our method on a genome with 99 windows of 400 loci each
produced the first plot in Figure \ref{fig:test_missing}.
These matrices are chosen so that the top two eigenvalues $\Sigma$ are the same (both 50.1),
and so the ordering of the top two PCs is arbitrary.
If our method was sensitive to PC ordering,
then half the windows in each region that have one ordering would cluster with each other,
separate from the other half. \llabel{ll:pc_switching_test}

We then marked each genotype in the first half of the chromosome
as missing, independently, with probability $1/2$ and ran our method again,
producing the second plot of Figure \ref{fig:test_missing}. \llabel{ll:test_missing}
If our method was influenced by missing data,
we would expect the first half of the chromosome to separate from the second
in the MDS plot.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
    \end{center}
    \caption{      
        PCA plots for the three sets of genomic windows colored in \Figure \ref{fig:mds_allchr},
        on each chromosome arm of \textit{Drosophila melanogaster}.
        In all plots, each point represents a sample. 
        The first column shows the combined PCA plot for windows 
        whose points are colored green in \Figure \ref{fig:mds_allchr}; 
        the second is for orange windows; 
        and the third is for purple windows.
         \label{fig:pca_by_pop}
    }
\end{figure}

\begin{figure}
    \begin{center}
        \includegraphics[width=1\textwidth]{Fig1_allchr_Together_MDS_plot_PCAk5_win103}
    \end{center}
    \caption{      
        MDS visualizations for each chromosome arm of \textit{Drosophila melanogaster},
        as in \Figure \ref{fig:mds_allchr},
        except that the method was run using five PCs ($k=5$) instead of two.
         \label{fig:dmel_k5}
    }
\end{figure}

>>>>

Population structure leads to systematic patterns in measures of mean relatedness
between individuals in large genomic datasets,
which are often discovered and visualized using dimension reduction techniques such as principal component analysis (PCA).
Mean relatedness
is an average of the relationships across locus-specific genealogical trees,
which can be strongly affected on intermediate genomic scales by linked selection and other factors.
We show how to use local principal components analysis to describe this 
meso-scale heterogeneity in patterns of relatedness,
and apply the method to genomic data from three species,
finding in each that the effect of population structure 
can vary substantially across only a few megabases.
In a global human dataset, localized heterogeneity is likely explained by polymorphic chromosomal inversions.
In a range-wide dataset of \textit{Medicago truncatula},
factors that produce heterogeneity are shared between chromosomes,
correlate with local gene density,
and may be caused by linked selection, such as background selection or local adaptation.
In a dataset of primarily African \textit{Drosophila melanogaster}, large-scale heterogeneity across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection,
and after removing samples carrying inversions, remaining heterogeneity is correlated with recombination rate and gene density,
again suggesting a role for linked selection.
% The method is implemented as an R package, called lostruct.
The visualization method provides a flexible new way to discover biological drivers of genetic variation,
and its application to data highlights the strong effects that linked selection and chromosomal inversions
can have on observed patterns of genetic variation.

% Article summary
% Principal component analysis is often used to describe overall "population structure", or patterns of relatedness arising from past demographic history, among a set of genomes.  In this paper, we describe how the patterns uncovered by PCA differ at a local scale along the genome.  We find that these patterns can be driven strongly by chromosomal inversions (and perhaps also regions of low recombination), and suggest that the strong remaining heterogeneity may be caused by the past action of linked selection.

@@

We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
in the space defined by the top principal components.
However, 
we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
since rotations or reflections of these imply identical patterns of relatedness.
Instead, we compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs,
which is invariant under reordering of the PCs, \revpoint{1}{2}
reflections, and rotations and yet contains all other information about the PCs.
(For results shown here, we use $k=2$.)
Furthermore, to remove the effect of artifacts such as mutation rate variation,
we also rescale each approximate covariance matrix to be of similar size
(precisely, so that the underlying data matrix has trace norm equal to one).

To do this, define the $N \times k$ matrix $V(i)$ so that $V(i)_{\cdot \ell}$, 
the $\ell^\text{th}$ column of $V(i)$,
is equal to the $\ell^\text{th}$ principal component of the $i^\text{th}$ window,
multiplied by $( \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} )^{1/2}$,
where $\lambda_{\ell i}$ is the $\ell^\text{th}$ eigenvalue of the genetic covariance matrix.
% \begin{equation}
%     \begin{aligned}
%         V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} ,
%         \qquad
%         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} 
%     \end{aligned}
% \end{equation}
Then, the rescaled, rank $k$ approximate covariance matrix for the $i^\text{th}$ window is
\begin{align} \label{eqn:est_cov}
    M(i) &= \sum_{\ell=1}^k V(i)_{\cdot \ell} V(i)_{\cdot \ell}^T .
\end{align}
% \begin{align}
%     M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
% \end{align}
To measure the similarity of patterns of relatedness for the $i^\text{th}$ window and $j^\text{th}$ window,
we then use
Euclidean distance $D_{ij}$ between the matrices $M(i)$ and $M(j)$,
defined by
$D_{ij}^2 = \sum_{k\ell} ( M(i)_{k,\ell} - M(j)_{k,\ell} )^2$.

@@

We tested the method using two types of simulation.
First, to verify expected behavior, we simulated ``genomes'' as an independent sequence
of correlated Gaussian ``genotypes'', using a different covariance matrix
in the first quarter, middle half, and last quarter of the chromosome.
The details of the simulation,
also designed to detect sensitivity to PC switching,
are given in Appendix \ref{apx:gaussian_sims}.
To verify robustness to missing data,
we ran the method after randomly dropping 50\% of the genotypes in the first half of the genome;
if the method is misled by missing data, then it will distinguish the two halves of the chromosome
rather than the segments having different covariance matrices.

To provide a realistic test,
we next used forwards-time, individual-based simulations, implemented using SLiM v3 \citep{haller2017flexible},
which are described in detail in Appendix \ref{apx:slim_sims}.
To provide realistic population structure for PCA to identify,
each simulation had at least 5,000 diploid individuals,   
living across a continuous square range,
with Gaussian dispersal and local density-dependent competition.
Each genome was modeled on human chromosome 7, which is $1.54 \times 10^8$bp long,
with an overall recombination rate of 1.6785 crossovers per chromosome per generation.
To improve speed, we used tskit \citep{kelleher2018efficient} to record tree sequences 
in SLiM \citep{haller2018treesequence}
and to add neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
Most simulations were neutral, but we also included linked selection, of two types.
First, % run_031486
we introduced selected mutations into two regions, 
which extended from 1/3 to 1/2 and from 5/6 to the end of the genome respectively.
These had selection coefficients from a Gamma distribution with shape 2 and mean 0.005
at a rate of $10^{-10}$ per bp, that were either beneficial (with probability $1/30$)
or deleterious (otherwise).
Second, % run_016913
to roughly model a recent expansion followed by local adaptation,
we introduced mutations in the same manner as above,
except that mutations were no longer unconditionally deleterious or beneficial:
each selection coefficient was multiplied by a factor depending on the spatial location
of the individual being evaluated, varying linearly from -1 at the left side of the range
to +1 at the right edge.
In all simulations, genome-wide PCA displayed a map of the population range, as expected.

@@

Simple non-population-based simulations with Gaussian ``genotypes''
showed that the method performs as expected, clearly separating regions of the genome
with different underlying covariance matrices
without being affected by extreme differences in amount of missing data
(Supplemental Figure \ref{sfig:test_missing}).
This simulation also verifies insensitivity to ordering of top PCs,
since it was performed using a covariance matrix with the top two eigenvalues equal,  
so that the order of empirical eigenvectors (PCs) switches randomly.

Individual-based simulations using SLiM \citep{haller2017flexible} allowed us to test the effects
of recombination and mutation rate variation, as well as linked selection.
As expected, varying recombination rate stepwise by a factor of 64
did not induce patterns in the MDS visualizations correlated with recombination rate
(Supplemental Figure \ref{sfig:slim_sims1}).
\llabel{ll:test_recomb_rate}
Since varying mutation rate with a fixed recombination map
is equivalent to varying the recombination map and remapping windows,
this also indicates that the method is not misled by variation in mutation rate.
On the other hand,
a recombination map with hotspots
(the HapMap human female map for chromosome 7 \citep{hapmap2007second})
induced outliers at long regions of low recombination rate
(also as expected). \llabel{ll:hotspot_recomb}

@@

We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
The first column of \Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
of the matrix of dissimilarities between genomic windows:
in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
For each chromosome arm, the MDS visualization roughly resembles a triangle,
sometimes with additional points.
Since the relative position of each window in this plot shows the similarity between windows, 
this suggests that there are at least three extreme manifestations of population structure 
typified by windows found in the ``corners'' of the figure,
and that other windows' patterns of relatedness may be a mixture of those extremes. 
The next two columns of \Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
plotted against the window's position along the genome,
to show how the plot of the first column is laid out along the genome.
The patterns did not depend on the number of PCs used
(see \Figure \ref{sfig:dmel_k5} for the same plot with $k=5$ PCs), \revpoint{1}{3}
and are only weakly correlated with variation in missingness 
(see \Figure \ref{sfig:dpgp_missingness}). \revpoint{2}{4}

@@

To help visualize how clustered windows with similar patterns of relatedness are along each chromosome arm, 
we selected three ``extreme'' windows in the MDS plot
and the 5\% of windows that are closest to it in the MDS coordinates,
then highlighted these windows' positions along the genome,
and created PCA plots for the windows, combined.
Representative plots are shown for three groups of windows on each chromosome arm
in \Figure \ref{fig:mds_allchr} (groups are shown in color),
and in Supplemental \Figure \ref{sfig:pca_by_pop} (PCA plots).
The latter plots are quite different,
showing that genomic windows in different regions of the MDS plot 
indeed show quite different patterns of relatedness.

The most striking variation in patterns of relatedness turns out to be explained by
several large inversions that are polymorphic in these samples, 
discussed in \citet{corbett2012population} and \citet{langley2012genomic}.
% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
To depict this, \Figure \ref{fig:pca_by_inversion} shows
the PCA plots in Supplemental \Figure \ref{sfig:pca_by_pop} recolored by the orientation of the inversion for each sample.
Taking chromosome arm 2L as an example,
the two regions of similar, extreme patterns of relatedness
shown in green in the first row of \Figure \ref{fig:mds_allchr}
lie directly around the breakpoints of the inversion In(2L)t,
and the PCA plots in the first rows of \Figure \ref{fig:pca_by_inversion}
shows that patterns of relatedness here are mostly determined by inversion orientation.
The regions shown in purple on chromosome 2L lie near the centromere,
and have patterns of relatedness reflective of two axes of variation,
seen in \Figure \ref{fig:pca_by_inversion} and Supplemental \Figure \ref{sfig:pca_by_pop},
which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
(see \citet{lack2015drosophila} for more about admixture in this sample).
The regions shown in orange on chromosome 2L mostly lie inside the inversion,
and show patterns of relatedness that are a mixture between the other two,
as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
Similar results are found in other chromosome arms,
albeit complicated by the coexistence of more than one polymorphic inversion;
however, each breakpoint visibly affects patterns in the MDS coordinates
(see vertical lines in \Figure \ref{fig:mds_allchr}).

@@

To see how patterns of relatedness vary in the absence of polymorphic inversions,
we performed the same analyses after removing, for each chromosome arm,
any samples carrying inversions on that arm.
In the result, shown in \Figure \ref{fig:drosophila_recomb_rate} and
Supplemental \Figure \ref{sfig:mds_allchr_noinversions},
the striking peaks associated with inversion breakpoints are gone,
and previously smaller-scale variation now dominates the MDS visualization.
For instance, the majority of the variation along 3L in \Figure \ref{fig:mds_allchr}
is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
there is also a relatively small dip on the right end of the arm (near the centromere).
In contrast, Supplemental \Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
remaining structure is dominated by the dip near the centromere.
Without inversions, variation in patterns of relatedness shown in the MDS plots
follows similar patterns to that previously seen in \textit{D.~melanogaster} recombination rate and diversity \citep{langley2012genomic,mackay2012drosophila}.
Indeed, correlations between the recombination rate in each window and the position on the first MDS coordinate are highly significant
(Spearman's $\rho=0.54$, $p<2 \times 10^{-16}$; \Figures \ref{fig:drosophila_recomb_rate} and \ref{sfig:drosophila_gene_density_correlations}).
This is consistent with the hypothesis that variation
is due to selection, since the strength of linked selection increases with local gene density, 
measured in units of recombination distance.
The number of genes -- measured as the number of transcription start and end sites within each window --
was not significantly correlated with MDS coordinate ($p=0.22$).

@@

The results were highly consistent across window sizes, window types (SNPs or bp),
and number of PCs, as shown in Table \ref{tab:param_cors}.
\revpoint{1}{4} \revpoint{1}{5}
% We also found nearly identical results when choosing shorter windows of 1,000 SNPs
% (Supplemental Figure \ref{sfig:mds_medicago_allchr_1e3}; \revpoint{1}{4}
% or choosing windows of equal length in base pairs rather than SNPs. \revpoint{1}{5}
% Similarly, the results were not substantially changed
% when using weighted PCA to downweight the large group of Tunisian samples.

\begin{figure}
    \begin{center}
       \includegraphics{Fig6_Together_MDS_plot_chr3_final}
       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
    \end{center}
    \caption{
        \textbf{(top)}
       MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3,
       with corresponding PCA plots.
       Each point in the plot represents a window;
       the structure revealed by the MDS plot is strongly clustered along the chromosome,
       with windows in the upper-right corner of the MDS plot (colored red) clustered around the centromere,
       windows in the upper-left corner (purple) furthest from the centromere,
       and the remaining corner (green) intermediate.
       Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:mds_medicago_allchr}.
       \textbf{(bottom)}
        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in the top figure.
        Each point corresponds to a sample, colored by country of origin.
        Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:pca_peaks_medicago_allchr}.
       \label{fig:mds12_medicago}
    }
\end{figure}

@@

\paragraph{Chromosomal inversions}
A major driver of variation in patterns of relatedness in two datasets we examined seems to be inversions.
This may be common,
but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
however, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
or simply regions of reduced recombination
without additional lines of evidence.  \llabel{ll:inversion_caveat}
Inversions show up in our method because across the inverted region,
most gene trees share a common split that dates back to the origin of the inversion.
However, in many applications, inversions are a nuisance.
For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
by regressing out the effect of linked SNPs on each other.
Removing samples with the less common orientation of each inversion reduced,
but did not eliminate, the signal of inversions
seen in the \textit{Drosophila melanogaster} dataset,
demonstrating that the genomic effects of transiently polymorphic inversions
may outlast the inversions themselves.

@@

Another possible contributor is recent admixture between previously separated populations,
the effects of which were not uniform across the genome due to selection. \revpoint{2}{6}
For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
is due to selection against Neanderthal genes, leading to greater introgression in regions of lower gene density
\citep{harris2016genetic,juric2016strength}.
African \textit{Drosophila melanogaster} are thought to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
if selection regularly favors genes from one origin,
this could lead to substantial variation in patterns of relatedness correlated with local gene density.

There has been substantial debate over the relative impacts of different forms of selection
(e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining,stankowski2018tempo}). \revpoint{2}{16}
These have been difficult to disentangle in part because for the most part
theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
It may be possible to design more powerful statistics that make stronger use of spatial information.

@@

\paragraph{Parameter choices}
There are several choices in the method that may in principle affect the results.
As with whole-genome PCA,
the choice of samples is important,
as variation not strongly represented in the sample will not be discovered.
The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
but downweighting may be a better option that does not discard data.
Next, the choice of window size may be important,
although in our applications results were not sensitive to this.
Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in \Figure \ref{fig:diagram}),
along with the method used to discover common structure,
will affect results.
We used MDS, applied to either each chromosome separately or to the entire genome;
for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosomes,
but genome-wide, their signal is obscured by the numerous other signals of comparable strength.

@@

We first created ``genotypes'' in this way with fifty individuals from each of the three populations;
running our method on a genome with 99 windows of 400 loci each
produced the first plot in Figure \ref{sfig:test_missing}.
These matrices are chosen so that the top two eigenvalues $\Sigma$ are the same (both 50.1),
and so the ordering of the top two PCs is arbitrary.
If our method was sensitive to PC ordering,
then half the windows in each region that have one ordering would cluster with each other,
separate from the other half. \llabel{ll:pc_switching_test}

We then marked each genotype in the first half of the chromosome
as missing, independently, with probability $1/2$ and ran our method again,
producing the second plot of Figure \ref{sfig:test_missing}. \llabel{ll:test_missing}
If our method was influenced by missing data,
we would expect the first half of the chromosome to separate from the second
in the MDS plot.

@@
